---
title: "vertexR: Turning Your Ontology into a System Graph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vertexR: Turning Your Ontology into a System Graph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What problem does this solve?

Data in relational tables is disconnected. You can query airports, airlines,
and routes individually, but the relationships between them -- which airlines
operate which routes from which airports -- live in join keys that are
invisible until you write SQL. When something goes wrong (an airport nears
capacity, a route gets cancelled), understanding the cascade of effects
requires mentally assembling the graph yourself.

**vertexR turns an ontology specification into an interconnected system
graph** so you can see cause-and-effect, not just rows. It materializes
every object instance as a node and every link instance as an edge,
then provides tools to explore, alert on, simulate, and optimize across
the whole system.

## Building a system graph

We'll use an aviation domain as a running example. First, set up a bundle
describing our ontology and populate a DuckDB with sample data.

```{r build-graph}
library(ontologySpecR)
library(vertexR)
library(dplyr, warn.conflicts = FALSE)

# -- Define the ontology bundle --
b <- bundle(
 bundle_id = "aviation-demo",
  bundle_version = "0.1.0",
  objects = list(
    object_type(
      id = "Airport",
      properties = list(
        property_def("airport_id", "string", nullable = FALSE),
        property_def("name", "string"),
        property_def("country", "string"),
        property_def("capacity", "integer"),
        property_def("utilization", "number"),
        property_def("latitude", "number"),
        property_def("longitude", "number")
      ),
      primary_key = "airport_id",
      source_kind = "table",
      source_table = "airports"
    ),
    object_type(
      id = "Airline",
      properties = list(
        property_def("airline_id", "string", nullable = FALSE),
        property_def("name", "string"),
        property_def("country", "string"),
        property_def("active", "boolean")
      ),
      primary_key = "airline_id",
      source_kind = "table",
      source_table = "airlines"
    ),
    object_type(
      id = "FlightRoute",
      properties = list(
        property_def("route_id", "string", nullable = FALSE),
        property_def("origin_id", "string"),
        property_def("destination_id", "string"),
        property_def("airline_id", "string"),
        property_def("stops", "integer"),
        property_def("equipment", "string")
      ),
      primary_key = "route_id",
      source_kind = "table",
      source_table = "routes"
    )
  ),
  links = list(
    link_type(
      id = "RouteOrigin",
      from = "FlightRoute", to = "Airport",
      cardinality = "many-to-one", directed = TRUE,
      join_from_keys = "origin_id", join_to_keys = "airport_id"
    ),
    link_type(
      id = "RouteDestination",
      from = "FlightRoute", to = "Airport",
      cardinality = "many-to-one", directed = TRUE,
      join_from_keys = "destination_id", join_to_keys = "airport_id"
    ),
    link_type(
      id = "RouteOperator",
      from = "FlightRoute", to = "Airline",
      cardinality = "many-to-one", directed = TRUE,
      join_from_keys = "airline_id", join_to_keys = "airline_id"
    )
  )
)

# -- Populate a DuckDB with sample data --
con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbWriteTable(con, "airports", data.frame(
  airport_id  = c("DUB", "JFK", "LHR", "CDG"),
  name        = c("Dublin", "JFK", "Heathrow", "Charles de Gaulle"),
  country     = c("Ireland", "USA", "UK", "France"),
  capacity    = c(30L, 100L, 80L, 70L),
  utilization = c(0.95, 0.6, 0.85, 0.5),
  latitude    = c(53.42, 40.64, 51.47, 49.01),
  longitude   = c(-6.27, -73.78, -0.46, 2.55),
  stringsAsFactors = FALSE
))
DBI::dbWriteTable(con, "airlines", data.frame(
  airline_id = c("EI", "AA", "BA"),
  name       = c("Aer Lingus", "American Airlines", "British Airways"),
  country    = c("Ireland", "USA", "UK"),
  active     = c(TRUE, TRUE, TRUE),
  stringsAsFactors = FALSE
))
DBI::dbWriteTable(con, "routes", data.frame(
  route_id       = c("R1", "R2", "R3", "R4", "R5"),
  origin_id      = c("DUB", "JFK", "LHR", "DUB", "CDG"),
  destination_id = c("JFK", "DUB", "JFK", "LHR", "LHR"),
  airline_id     = c("EI", "EI", "BA", "EI", "BA"),
  stops          = c(0L, 0L, 0L, 0L, 1L),
  equipment      = c("A330", "A330", "777", "A320", "A320"),
  stringsAsFactors = FALSE
))

# -- Build the system graph --
g <- vertex_graph(b, con)
g
```

One call turns three database tables and three link definitions into a
unified graph. Each node carries its `.object_type` and `.node_id`, plus
all property columns from the source table.

## Exploring the graph

### Node and edge accessors

```{r explore-nodes}
vx_nodes(g)
```

```{r explore-edges}
vx_edges(g)
```

### Subgraph by object type

Pull out just the airports:

```{r subgraph}
airport_graph <- vx_subgraph(g, "Airport")
vx_nodes(airport_graph)
```

### Search-around (neighbors)

Starting from Dublin, find everything within 1 hop:

```{r neighbors}
dub_nbrs <- vx_neighbors(g, "DUB", depth = 1)
vx_nodes(dub_nbrs)
```

At depth 2, we reach further:

```{r neighbors-2}
dub_nbrs2 <- vx_neighbors(g, "DUB", depth = 2)
vx_nodes(dub_nbrs2)
```

### Shortest path

Find the path between Dublin and JFK:

```{r shortest-path}
vx_shortest_path(g, "DUB", "JFK")
```

### Connected components

Identify which parts of the graph are connected:

```{r components}
vx_connected_components(g, mode = "weak")
```

### Cycle detection

Check if the graph contains directed cycles (useful for validation):

```{r cycles}
vx_cycle_detect(g)
```

Our aviation graph is a DAG (routes point to airports/airlines but not
vice versa), so this returns `FALSE`.

## Setting up alerts

Alerts let you define conditions on the graph and evaluate them
programmatically. Each rule is a function that receives the graph and
returns the `.node_id` values of flagged nodes.

```{r alerts}
rule <- vx_alert_rule(
  id = "over_capacity",
  description = "Airports over 90% utilization",
  condition = function(g) {
    nodes <- vx_nodes(g)
    airports <- nodes[nodes$.object_type == "Airport", ]
    airports[airports$utilization > 0.9, ]$.node_id
  },
  severity = "critical"
)

alerts <- vx_evaluate_alerts(g, list(rule))
alerts
```

Dublin (DUB) is flagged -- it's at 95% utilization.

### Propagating risk

Which nodes are downstream of Dublin? If DUB has a problem, which routes
and connected airports are affected?

```{r propagate}
vx_propagate_risk(g, "DUB", depth = 2)
```

## What-if scenarios

Scenarios let you modify the graph and see what happens without touching
the database. What if Dublin airport closes?

```{r scenario}
scenario <- vx_scenario(g) |>
  vx_remove_node("DUB")

g_closed <- vx_apply_scenario(scenario)
g_closed
```

### Comparing original vs. scenario

```{r compare}
diff <- vx_compare(g, g_closed)
diff
```

The comparison shows that DUB was removed along with its incident edges --
the three routes that originated from or went to Dublin.

You can also modify properties:

```{r scenario-modify}
scenario2 <- vx_scenario(g) |>
  vx_modify_node("DUB", capacity = 50L, utilization = 0.6)

g2 <- vx_apply_scenario(scenario2)
vx_compare(g, g2)
```

Now DUB's capacity has increased and utilization has dropped -- we can
re-evaluate alerts on the modified graph:

```{r scenario-alerts}
vx_evaluate_alerts(g2, list(rule))
```

No alerts -- the scenario resolved the over-capacity issue.

## Visualization

vertexR provides interactive graph visualization via visNetwork (with a
ggraph fallback for static plots).

```{r plot, eval = requireNamespace("visNetwork", quietly = TRUE)}
vx_plot(g,
  color_by = ".object_type",
  size_by = "capacity",
  highlight = "DUB",
  label_by = ".node_id"
)
```

You can also visualize scenario diffs:

```{r plot-scenario, eval = requireNamespace("visNetwork", quietly = TRUE)}
scenario_viz <- vx_scenario(g) |>
  vx_remove_node("DUB")
vx_plot_scenario(g, scenario_viz)
```

### Concept integration

If you use `conceptR` to evaluate concepts on your objects, you can join
those results onto the graph for visualization:
```{r concept-viz, eval = FALSE}
# concept_results from conceptR::cpt_evaluate()
concept_results <- data.frame(
  .node_id = c("DUB", "JFK", "LHR", "CDG"),
  busy_airport = c(TRUE, FALSE, TRUE, FALSE)
)

g_concepts <- vx_color_by_concept(g, concept_results)
vx_plot(g_concepts, color_by = "busy_airport")
```

## Memory management

By default, `vertex_graph()` guards against loading excessively large graphs
into memory. The `max_nodes` parameter (default 50,000) will abort with a
clear message if the total node count exceeds the limit:

```{r max-nodes, eval = FALSE}
# This would fail if the database had > 50k total rows:
g <- vertex_graph(b, con, max_nodes = 50000L)

# Disable the guard for very large graphs (use with caution):
g <- vertex_graph(b, con, max_nodes = Inf)
```

For large-scale queries, consider using `objectSetsR` for lazy evaluation
before materializing a smaller subgraph with vertexR. Alternatively, use
the DuckPGQ backend described below.

## DuckPGQ Backend (Large Graphs)

For graphs too large to fit in memory, vertexR provides a complementary
backend that executes graph queries inside DuckDB using the DuckPGQ extension.
Instead of materializing nodes and edges in R, queries run as SQL/PGQ
statements and return only the results you need.
### When to use which backend

| Backend | Use Case | Memory |
|---------|----------|--------|
| `vertex_graph()` + igraph | Small-to-medium graphs, algorithm-heavy work | In-memory |
| `vx_pgq_*()` + DuckPGQ | Large graphs, traversal queries | In-database |

The recommended pattern: use `vx_pgq_*` for traversal on the full dataset,
then pull a subgraph into `vertex_graph()` for scenario analysis and
optimization.

### Setting up DuckPGQ

First, check if DuckPGQ is available:

```{r pgq-setup, eval = FALSE}
# Check availability
vx_pgq_available(con)

# Create property graph from bundle
vx_pgq_setup(b, con)
graph_name <- vx_pgq_graph_name(b)  # "aviation_demo"
```

The setup creates a DuckPGQ property graph with vertex tables for each
object type and edge tables for each link type.

### Querying neighbors

Find what's connected to a node without loading the full graph:

```{r pgq-neighbors, eval = FALSE}
# Find routes originating from DUB (executes in DuckDB)
vx_pgq_neighbors(con, graph_name,
  source_type = "Airport",
  source_ids = c("DUB"),
  link_type = "RouteOrigin",
  target_type = "FlightRoute",
  depth = 1,
  direction = "backward"  # routes point TO airports
)
```

### Finding shortest paths

```{r pgq-path, eval = FALSE}
# Find path between two airports
vx_pgq_shortest_path(con, graph_name,
  from_type = "Airport", from_id = "DUB",
  to_type = "Airport", to_id = "JFK",
  max_hops = 5
)
```

### Investigation patterns (ICIJ-style)

vertexR includes pre-built patterns for common investigative queries,
inspired by the ICIJ's offshore leaks analysis:

```{r pgq-patterns, eval = FALSE}
# List available patterns
vx_pgq_list_patterns()

# Find entities sharing a common officer
vx_pgq_pattern(con, graph_name, "common_officer")

# Find officers spanning multiple jurisdictions
vx_pgq_pattern(con, graph_name, "multi_jurisdiction_officer",
  params = list(min_jurisdictions = 3))

# Find intermediaries connecting many entities
vx_pgq_pattern(con, graph_name, "intermediary_hub",
  params = list(min_connections = 10))
```

### Extracting subgraphs for algorithm work

After identifying interesting nodes with DuckPGQ queries, pull them into
an in-memory graph for detailed analysis:

```{r pgq-subgraph, eval = FALSE}
# Extract subgraph around specific nodes
g <- vx_pgq_subgraph(con, b, graph_name,
  seed_ids = c("DUB", "JFK"),
  seed_type = "Airport",
  depth = 2
)

# Now use in-memory functions
vx_scenario(g) |>
  vx_remove_node("DUB") |>
  vx_apply_scenario()
```

This two-phase approach lets you work with arbitrarily large graphs:
DuckPGQ handles the broad traversal, then vertexR's in-memory engine
handles the detailed what-if analysis.

## How it fits the ecosystem

```
                     +-------------------+
                     |   Your R session  |
                     +-------------------+
                              |
                     +--------v--------+
                     |     vertexR     |
                     | (graph, alerts, |
                     |  scenarios, opt)|
                     +--------+--------+
                              |
      +-----------+-----------+-----------+
      |           |                       |
+-----v-----+ +---v---+          +--------v--------+
| vertex_   | |vx_pgq_|          |       DBI       |
| graph()   | |  *()  |          | (DuckDB, PG,    |
| (igraph)  | |(PGQ)  |          |  SQLite, ...)   |
+-----------+ +---+---+          +-----------------+
                  |                       |
                  +--------+------+-------+
                           |      |
                 +---------v--+ +-v---------+
                 |ontologySpecR| |  DuckPGQ  |
                 |(bundle,     | |(optional) |
                 | types)      | +-----------+
                 +-------------+
```

- **ontologySpecR** provides the type definitions (bundle, object types,
  link types). vertexR reads these to know which tables to query and how
  to join them.
- **DBI** is the database abstraction. vertexR works with any DBI backend --
  DuckDB for local analytics, PostgreSQL for production, etc.
- **DuckPGQ** is an optional DuckDB extension for graph queries. When
  installed, vertexR's `vx_pgq_*` functions execute traversals directly
  in the database without materializing the full graph in R.
- **objectSetsR** is an optional companion for lazy object set queries.
  vertexR builds its own graph construction to avoid known issues in
  objectSetsR's primary key handling.
- **actionTypesR** is an optional companion. After a simulation identifies
  a needed intervention, you could execute it as an action via actionTypesR.
- **conceptR** is an optional companion for concept evaluation. Use
  `vx_color_by_concept()` to visualize concept membership on the graph.

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
